---
title: "Alina's PhD Data Analysis Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(tidyr)
```

## Setting the Stage

1. A set of 13 different tables provided
2. Objective: To determine relationships/correlations between different variables of the table.  


## EDA

1. Reduce Variables and clean up names
2. Clean up missing Values
3. Perform Basic Visualization
4. Remove Outliers - 3 different approaches
4. Filter out outlier values based on statistical approaches
5. Hypothesis to check?

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}
d1 <- read.csv("~/R/Rtuts/Data/C57BL6_476_vs_C57BL6_ctl.csv")
head(d1)
```
```{r}
str(d1)
```

Removing unneccesary columns to reduce the table!

```{r}
d2 <- select(d1, -lfcSE, -stat, -symbol)
head(d2)
```

```{r}
str(d2)
```

 Rename the Variable names into more readable ones!
```{r}
d3 <- d2 %>% rename(CPM_476_R1 = CPM_.C57BL6_476_R1.) %>% rename(CPM_476_R2 = CPM_.C57BL6_476_R2.) %>% rename(CPM_ctl_R1 = CPM_.C57BL6_ctl_R1.)  %>% rename(CPM_ctl_R2 = CPM_.C57BL6_ctl_R2.)
head(d3)
```




```{r}
summary(d3)
```


### Question1 :

What does the value 0.0 in the min for CPM indicate? Is it a real value or worthy to be discarded?

### Checking Missing Values

```{r}
glimpse(d3)
```


###
```{r}
d3 %>% summarise(c1 = sum(is.na(CPM_476_R1)),
                 c2 = sum(is.na(CPM_476_R2)),
                 c3 = sum(is.na(CPM_ctl_R1)),
                 c4 = sum(is.na(CPM_ctl_R1))
                 )
```

Therefore, there are no missing values in the table.


## Basic Visualization to get a sneak peak at Data!

```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange, y = baseMean)) + geom_point(alpha = 0.1)

```




```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange, y = pvalue)) + geom_point(alpha = 0.1)
```


```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange, y = padj)) + geom_point(alpha = 0.1)
```



```{r}
ggplot(data = d3, mapping = aes(x = CPM_476_R1, y = CPM_ctl_R1)) + geom_point(alpha = 0.1)
```

```{r}
ggplot(data = d3, mapping = aes(x = CPM_476_R2, y = CPM_ctl_R2)) + geom_point(alpha = 0.1)
```


```{r}
ggplot(data = d3, mapping = aes(x = CPM_ctl_R1, y = CPM_ctl_R2)) + geom_point(alpha = 0.1)
```



```{r}
ggplot(data = d3, mapping = aes(x = CPM_476_R1, y = CPM_476_R2)) + geom_point(alpha = 0.1)
```


### Hence, there exists major issue with "Outliers" in data, making the data skewed on the wrong end!

## Removing Outliers


Calculating Standard Deviation from current table gives misplaced picture. Hence, remove outliers and then calculate standard deviation to place meaningful thresholds!

The best way to look for outliers in two numeric variables is using a scatter plot.

```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange, y = pvalue)) + geom_boxplot() + geom_point(alpha = 0.1, color = "blue")
```


```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange, y = pvalue)) + geom_boxplot() + geom_point(alpha = 0.009, color = "blue")
```
```{r}
ggplot(data = d3, mapping = aes(x = CPM_476_R1)) + geom_dotplot(binwidth = 400, color = "brown") 
```


```{r}
ggplot(data = d3, mapping = aes(x = baseMean)) + geom_dotplot(binwidth = 8000, color = "green")
```


```{r}
ggplot(data = d3, mapping = aes(x = padj)) + geom_histogram( color = "green")
```


```{r}
ggplot(data = d3, mapping = aes(x = log2FoldChange)) + geom_histogram(binwidth = 0.15, color = "blue")
```


### Question 2: 
The values of other variables are derived from a (some particular) transformation of CPM values? If yes, then filtering outliers from CPM will also remove values from other variables and could lead elimination of certain genes. Would that create a problem?


### Question 3: 
No wrongly tabulated or mistakenly tabulated data? More an Art than Science!

### Question 4: Which approach to pick for removing outliers?
1. Hard cut as a threshold based on visualization.
2. Based on Z-scores.
3. Based on IQR.
4. Based on Spread and variability of data : 1sigma, 2 sigma and 3 sigman spread.





