---
title: 'DGE Analysis Notebook - 755 vs LI'
output:
  html_document: 
    toc: yes
    fig_width: 10
    fig_height: 12
    fig_caption: yes
    number_sections: yes
    toc_depth: 4
  pdf_document: 
    toc: yes
    number_sections: yes
---

```{r, message = FALSE, warning = FALSE}
suppressPackageStartupMessages(library("DESeq2"))
suppressPackageStartupMessages(library("pheatmap"))
suppressPackageStartupMessages(library("PoiClaClu"))
suppressPackageStartupMessages(library("RColorBrewer"))
suppressPackageStartupMessages(library('tidyverse'))
suppressPackageStartupMessages(library("PoiClaClu"))
suppressPackageStartupMessages(library("vsn"))
suppressPackageStartupMessages(library('EnhancedVolcano'))
suppressPackageStartupMessages(library('gplots'))
suppressPackageStartupMessages(library('org.Mm.eg.db'))
suppressPackageStartupMessages(library('stringr'))
suppressPackageStartupMessages(library("genefilter"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("glmpca"))
suppressPackageStartupMessages(library('org.Mm.eg.db'))
suppressPackageStartupMessages(library("AnnotationDbi"))
suppressPackageStartupMessages(library("apeglm"))
suppressPackageStartupMessages(library("ComplexHeatmap"))
suppressPackageStartupMessages(library("clusterProfiler"))
suppressPackageStartupMessages(library('ggrepel'))
suppressPackageStartupMessages(library('corrplot'))
suppressPackageStartupMessages(library("GO.db"))
suppressPackageStartupMessages(library('edgeR'))
suppressPackageStartupMessages(library('GOstats'))
suppressPackageStartupMessages(library('pathview'))
suppressPackageStartupMessages(library("gage"))
suppressPackageStartupMessages(library("gageData"))
suppressPackageStartupMessages(library('GOSemSim'))
suppressPackageStartupMessages(library('DOSE'))
suppressPackageStartupMessages(library('enrichplot'))
suppressPackageStartupMessages(library('ggnewscale'))
suppressPackageStartupMessages(library('glue'))
suppressPackageStartupMessages(library("ggupset"))
suppressPackageStartupMessages(library("SPIA"))
suppressPackageStartupMessages(library("stats"))
```

# Differential Gene Expression Analysis

## Creating metadata for the DGE Analysis

DESeq2 needs sample information (metadata) for performing DGE analysis.
Let's create the sample information

```{r}
# Read the csv file and change the column name. the samples.csv is a list of sample names, ie, the names of bam files.
sample_ID <- read.csv("~/R/Alina_RNAseq/samples.csv")
head(sample_ID)
```

```{r}
condition <- c("Infected", "Infected", "Infected", "Infected", "Infected", "Infected",
               "Infected", "Infected", "Infected", "Infected", "Infected", "Infected",
               "Infected", "Infected", "control", "control")
coldata <- data.frame(sample_ID, condition)
colnames(coldata) <- c('Sample_Name','condition') # change name of one of the columns
# The metadata can be found in a df called coldata!
head(coldata)
```

### Tidying up the names for plots later!

#### First from coldata

```{r}
# tidying up the names od samples in both columns that list of samples
#coldata$Samples <- str_remove_all(coldata$Sampls, pattern = "run6_trimmed_|_.bam|_S\\d\\d|_S\\d")
coldata$Sample_Name <- str_remove_all(coldata$Sample_Name,
                                  pattern = "run6_trimmed_|_.bam|_S\\d\\d|_S\\d" )
coldata$condition <- as.factor(coldata$condition)
```

### Changing the names of samples (as per Alina)

```{r}
coldata[coldata == '476_R1'] <- 'T'
coldata[coldata == '754_R1'] <- 'S54'
coldata[coldata == '755_R1'] <- 'S55'
coldata[coldata == '757_R1'] <- 'L57'
coldata[coldata == '758_R1'] <- 'A58'
coldata[coldata == '760_R1'] <- 'L60'
coldata[coldata == '761_R1'] <- 'S61'
coldata[coldata == '762_R1'] <- 'A62'
coldata[coldata == '763_R1'] <- 'L63'
coldata[coldata == '764_R1'] <- 'A64'
coldata[coldata == '765_R1'] <- 'S65'
coldata[coldata == '766_R1'] <- 'L66'
coldata[coldata == '768_R1'] <- 'A68'
coldata[coldata == '769_R1'] <- 'L69'
coldata[coldata == 'Ctrl1_R1'] <- 'C1'
coldata[coldata == 'Ctrl2_R2'] <- 'C2'
# convert column1 with sample names to row.names of coldata
rownames(coldata) <- coldata$Sample_Name
coldata
```

## Adding the groupings by Alina for further Metadata Information

```{r}
coldata$Epithelial_response <- c("LowInducer", "LowInducer", "HighInducer",
                                  "HighInducer", "LowInducer", "LowInducer",
                                  "HighInducer", "HighInducer", "LowInducer",
                                  "HighInducer", "LowInducer", "HighInducer",
                                  "LowInducer", "LowInducer", 'NR', 'NR')
coldata$clinical_outcome <- c('symptomatic', 'symptomatic', 'symptomatic',
                              'Lethal', 'asymptomatic', 'Lethal', 'symptomatic',
                              'asymptomatic', 'Lethal', 'symptomatic', 'symptomatic',
                              'Lethal', 'asymptomatic', 'Lethal', 'NR', 'NR')
coldata$microcolonies <- c('Low', 'Low', 'Low', 'High', 'Low', 'Low',
                           'High', 'High', 'Low', 'High', 'Low', 'High', 'Low',
                           'Low', 'NR', 'NR')
coldata$ER_microcolonies <- c("LI_LM", "LI_LM", "HI_LM", "HI_HM", "LI_LM", "LI_LM",
                              "HI_HM", "HI_HM", "LI_LM", "HI_HM", "LI_LM", "HI_HM",
                              "LI_LM", "LI_LM", 'NR', 'NR')
coldata$phylogenomic_lineage <- c("EPEC1", "EPEC10", "EPEC9", "EPEC9", "NC", "EPEC5",
                                  "EPEC8", "NC", "EPEC7", "NC", "EPEC2", "EPEC9",
                                  "EPEC2", "EPEC2", 'NR', 'NR')
coldata$phylogroup <- c("B2", "A", "B2", "B2", "B1", "A", "B2", "B2", "B1", "B2", "B1",
                        "B2", "B1", "B2", 'NR', 'NR')
coldata$Intimin_Type <- c("alpha", "ND", "lambda", "lambda", "epsilon", "epsilon",
                          "mu", "lambda", "beta", "kappa", "beta", "alpha", "beta",
                          "beta", 'NR', 'NR')
```

#### then fix Countsmatrix:

NOTE:

1.  From the manuals the countsData must be a numeric matrix
2.  It is IMPORTANT to keep the names of the genes in the rownames

```{r}
# Readin  countsmatrix
#countsmatrix <-as.matrix(read.csv("~/R/Rtuts/Data/Alina_EPEC_project/counts.csv"))
countsmatrix <- read.csv("~/R/Alina_RNAseq/newcounts.csv")
#countsmatrix <- as.data.frame(countsmatrix)
```

```{r}
## Removal of Gender Genes from ENSEMBL ID itself
countsmatrix <- countsmatrix %>% filter(countsmatrix$X != "ENSMUSG00000086503",
                                  countsmatrix$X != "ENSMUSG00000097571",
                                  countsmatrix$X != "ENSMUSG00000086370",
                                  countsmatrix$X != "ENSMUSG00000031329")
nrow(countsmatrix)
countsmatrix <- as.matrix(countsmatrix)
```

```{r}
#tidying up these names again
#colnames(countsmatrix) <- str_remove_all(colnames(countsmatrix), pattern = "run6_trimmed_|_.bam|_S\\d\\d|_S\\d")
rownames(countsmatrix) <- countsmatrix[,1] #converting first column of gene names into rownames, to be used for sanity check later
# It is IMPORTANT to keep the names of the genes in the rownames
countsmatrix <- subset(countsmatrix, select = - X)#dropping the X column
# the elements from Sample_Name from coldata must the the colnames of countsmatrix
colnames(countsmatrix) <- coldata$Sample_Name
# Display the column names
colnames(countsmatrix)
# Convert the countsmatrix elements to be of numeric type in order to be in correct format to be fed into DESeq2 functions
class(countsmatrix) <- "numeric"
```

# Calculating CPM Values

```{r}
# as DGEList
dge_755a <- DGEList(counts = countsmatrix)

dim(dge_755a)
colnames(dge_755a)
#dge_er$samples

## calculate norm. factors
nr <- calcNormFactors(dge_755a)

## get normalized counts
cpmvalues <- cpm(nr)
cpmvalues_d <- cpm.default(nr)
```

# Differential Gene Expression analysis using DESeq2

Now, construct DESeqDataSet for DGE analysis.

But before that, a sanity check : It is essential to have the name of
the columns in the count matrix in the same order as that in name of the
samples (rownames in coldata).

```{r}
all(rownames(coldata) %in% colnames(countsmatrix))
ncol(countsmatrix) == nrow(coldata)
dim(countsmatrix)
```

## Creating the DESeq Data set Object

```{r}
dds_755a <- DESeqDataSetFromMatrix(countData = countsmatrix,
                              colData = coldata, 
                              design = ~ ER_microcolonies)
nrow(dds_755a)
```

```{r}
# Function to save generic plots
saveplot <- function(plot,name ){
  # Function to save the plots
  ggsave(filename = 
           glue('~/R/Alina_RNAseq/755vsLI/{name}.png'),
       plot = plot,
       dpi = 300,
       width = 10,
       height = 10,
       units = "in")
}
```

## Exploratory Data Analysis and Visualization

### Pre-filtering the dataset

Our count matrix with our DESeqDataSet contains many rows with only
zeros, and additionally many rows with only a few fragments total. In
order to reduce the size of the object, and to increase the speed of our
functions, we can remove the rows that have no or nearly no information
about the amount of gene expression.

Applying the most minimal filtering rule: removing rows of the
DESeqDataSet that have no counts, or only a single count across all
samples. Additional weighting/filtering to improve power is applied at a
later step in the workflow.

```{r}
keep <- rowSums(counts(dds_755a)) > 1
dds_755a <- dds_755a[keep,]
nrow(dds_755a)
```

### The variance stabilizing transformation

## Applying VST transformation

```{r}
vsd <- vst(dds_755a, blind = FALSE)
head(assay(vsd), 3)
colData(vsd)
vsd_coldata <- colData(vsd)
```

```{r}
dds_755a <- estimateSizeFactors(dds_755a)
dds_755a
```

## Sample Distances

useful first step in an RNA-seq analysis is often to assess overall
similarity between samples:

1.  Which samples are similar to each other, which are different?
2.  Does this fit to the expectation from the experiment's design?

### Euclidean Distance between samples

dist to calculate the Euclidean distance between samples - useful for
ONLY normalized data. To ensure we have a roughly equal contribution
from all genes, we use it on the VST data.

```{r}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- vsd$Sample_Name
colnames(sampleDistMatrix) <- vsd$Sample_Name

colors <- colorRampPalette( rev(brewer.pal(9, "RdYlBu")) )(255)
(EuclideanDistanceHeatmap <- pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         main = "Sample-to-Sample Euclidean Distance of BL6 - 755A",
         col = colors))

```

```{r}
#saveplot(EuclideanDistanceHeatmap, "EuclideanDistanceHeatmap")
```

### Poisson Distance between Samples

```{r}
poisd <- PoissonDistance(t(counts(dds_755a))) # raw counts or unnormalised data
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- dds_755a$Sample_Name
colnames(samplePoisDistMatrix) <- dds_755a$Sample_Name

colors <- colorRampPalette( rev(brewer.pal(9, "RdYlBu")) )(255)
(poisson_dist_plot <- pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         main = "Sample-to-Sample Poisson Distance of BL6 - 755A",
         col = colors))
         
```

```{r}
#saveplot(poisson_dist_plot, "poisson_dist_plot")
```

# PCA Plot

## Calculating all PCA Values

```{r}
# calculate the variance for each gene
rv <- rowVars(assay(vsd))
ntop <- 500
# select the ntop genes by variance
select <- order(rv, decreasing = TRUE)[seq_len(min(ntop, length(rv)))]

# perform a PCA on the data in assay(x) for the selected genes
pca_755a <- prcomp(t(assay(vsd)[select,]))
summary(pca_755a)
# the contribution to the total variance for each component
#percentVar_755a <- (pca_755a$sdev^2 / sum( pca_755a$sdev^2 )) * 100
#percentVar_755a
```

## PCA Plot with VST Data

### Function for calculating percentvar

```{r}
percentvar_calculation <- function(pcaData_variable){
  # function to calculate percentvar for different variables
  percentvar_variable <- round(100 * attr(pcaData_variable, "percentVar"), digits = 3 )
  return(percentvar_variable)
}

savingFunction <- function(plotname, metadatacolumn){
  # Function to save the PCA plots
  ggsave(filename = 
           glue('~/R/Alina_RNAseq/755vsLI/PCAplot_HighVsLowInducer_{metadatacolumn}.png'),
       plot = plotname,
       dpi = 300,
       width = 10,
       height = 10,
       units = "in")
}
```

```{r}
pcaData_755a <- plotPCA(vsd, intgroup = c("ER_microcolonies",
                                                "Sample_Name"), 
                              returnData = TRUE)
percentVar_755a <- percentvar_calculation(pcaData_755a)
```

### Functions for Plot aethetics and saving PCA Plots

```{r}
color_values <- c("red", "red","red","red", "black","black","red","red", "red",
                  "red", "red","red","red", "red", "red" ,"blue")

# the basic set of common aesthetic settings for PCA plots, 
theme.my.own <- list(theme_bw() ,
                      geom_point(size = 3),
                      coord_fixed() ,
                      scale_y_continuous(breaks = seq(-20, 20, 5), 
                                         sec.axis = sec_axis(~. *1, 
                                                             labels = NULL,
                                                             breaks = NULL)) ,
                      scale_x_continuous(breaks = seq(-20, 20, 5), 
                                         sec.axis = sec_axis(~. *1, 
                                                             labels = NULL,
                                                             breaks = NULL)) ,
                      theme_classic() ,
                      geom_hline(yintercept = 0, color = "gray", size = 1) ,
                      geom_vline(xintercept = 0, color = "gray", size = 1) ,
                      theme(text = element_text(size = 15),
                            axis.text = element_text(size = 15),
                            #sec.axis = element_blank(),
                            legend.position = "right",
                            aspect.ratio = 1) ,
                      #geom_text(size = 4, hjust = 0, vjust = 0)
                      geom_text_repel(size = 5,min.segment.length = 0.5)
  )

```

```{r }
(PCAplot_vst <- ggplot(pcaData_755a,
                      aes(x = PC1,
                          y = PC2,
                          color = Sample_Name,
                          label = Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies") +
        scale_colour_manual(values = color_values) +
        theme.my.own )

savingFunction(PCAplot_vst, "ER_Microcolonies")
```

## PCA Plot for different groupings of metadata

### PCA for Microcolonies

```{r}
PCAdata_755a_MC <- plotPCA(vsd, intgroup = c("Sample_Name", "microcolonies"), returnData = TRUE)
percentVar_755a_MC <- percentvar_calculation(PCAdata_755a_MC)
```

```{r}
(PCAplot_MC <- ggplot(PCAdata_755a_MC,
                     aes(x = PC1, y = PC2,
                         color = microcolonies,
                         label= Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a_MC[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a_MC[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies755A - Microcolonies") +
        theme.my.own)

savingFunction(PCAplot_MC, "microcolonies")
```

### PCA for Clinical Outcome

```{r}
PCAdata_755a_CO <- plotPCA(vsd, 
                                 intgroup = c("Sample_Name", "clinical_outcome"), 
                                 returnData = TRUE)
percentVar_755a_CO <- percentvar_calculation(PCAdata_755a_CO)
```

```{r}
(PCAplot_CO <- ggplot(PCAdata_755a_CO,
                     aes(x = PC1, y = PC2,
                         color = clinical_outcome,
                         label = Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a_CO[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a_CO[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies755A - Clinical Outcome") +
        theme.my.own)

savingFunction(PCAplot_CO, "clinical_outcome")
```

### PCA for Phylogenomic Lineage

```{r}
PCAdata_755a_PL <- plotPCA(vsd, intgroup = c("Sample_Name", 
                                                   "phylogenomic_lineage"), 
                                 returnData = TRUE)
percentVar_755a_PL <- percentvar_calculation(PCAdata_755a_PL)
```

```{r}
(PCAplot_PL <- ggplot(PCAdata_755a_PL,
                       aes(x = PC1, y = PC2,
                           color = phylogenomic_lineage,
                           label = Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a_PL[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a_PL[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies755A - Phylogenomic Lineage") +
        theme.my.own)

savingFunction(PCAplot_PL, "phylogenomic_lineage")
```

### PCA for Phylogroup

```{r}
PCAdata_755a_PG <- plotPCA(vsd, intgroup = c("Sample_Name", "phylogroup"), returnData = TRUE)
percentVar_755a_PG <- percentvar_calculation(PCAdata_755a_PG)
```

```{r}
(PCAplot_PG <- ggplot(PCAdata_755a_PG,
                     aes(x = PC1, y = PC2,
                         color = phylogroup,
                         label= Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a_PG[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a_PG[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies755A - Phylogroup") +
        theme.my.own)

savingFunction(PCAplot_PG, "phylogroup")
```

### PCA for Intimin Group

```{r}
PCAdata_755a_IT <- plotPCA(vsd, intgroup = c("Sample_Name", "Intimin_Type"), returnData = TRUE)
head(PCAdata_755a_IT)
percentVar_755a_IT <- round(100 * attr(PCAdata_755a_IT, "percentVar"), digits = 4)
```

```{r}
(PCAplot_IT <- ggplot(PCAdata_755a_IT,
                     aes(x = PC1, y = PC2,
                         color = Intimin_Type,
                         label = Sample_Name)) +
        xlab(paste0("PC1: ", percentVar_755a_IT[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar_755a_IT[2], "% variance")) +
        ggtitle("PCA Plot - ER_Microcolonies755A - Intimin type") +
        theme.my.own)

savingFunction(PCAplot_IT, "Intimin_Type")
```

## Hierarchical Clustering

### applying rlog Transformation

```{r}
rld <- rlog(dds_755a, blind = FALSE)
head(assay(rld), 3)
```

```{r}
### Extract the rlog matrix from the object
rld_mat <- assay(rld) #assay() is function from the "SummarizedExperiment" package that was loaded when you loaded DESeq2
```

```{r}
### Compute pairwise correlation values
rld_cor <- cor(rld_mat)    ## cor() is a base R function
head(rld_cor)   ## check the output of cor(), make note of the rownames and colnames
```

```{r}
### Plot heatmap
heat.colors <- brewer.pal(6, "RdYlBu")
(Hclust_plot <- pheatmap(rld_cor, 
                        color = heat.colors,
                        main = 'Heirarchical Clustering of Samples - ER_Microcolonies755A - Correlation Matrix'
                        #filename = '~/R/Alina_RNAseq/EpithelialResponse/Hclust_plot.tiff'
                       ))
#Hclust_plot
```

```{r}
#saveplot(Hclust_plot, "Hclust_plot")
```

## DGE Results

### Running the differential expression pipeline

```{r}
dds1_755a <- DESeq(dds_755a)
#str(dds1)
```

### Building the results table

```{r}
res_755a <- results(dds1_755a,contrast = c("ER_microcolonies","HI_LM","LI_LM"))
head(res_755a, 30)
```

```{r}
summary(res_755a)
```

A Problem: there a number of rows with padj that has NA.

Reaons:

1.  If within a row, all samples have zero counts, the baseMean column
    will be zero, and the log2 fold change estimates, p value and
    adjusted p value will all be set to NA. ------\>\>\>\>\>\> checked
    and not true!

2.  If a row contains a sample with an extreme count outlier then the p
    value and adjusted p value will be set to NA. These outlier counts
    are detected by Cook's distance. Customization of this outlier
    filtering and description of functionality for replacement of
    outlier counts and refitting is described below. ------\>\>\>\>\>\>
    checked and not true!

3.  If a row is filtered by automatic independent filtering, for having
    a low mean normalized count, then only the adjusted p value will be
    set to NA. Description and customization of independent filtering is
    described below. ------\>\>\>\>\>\> could be true!

=====\>\>\>\>\>\> Most likely caused by automatic independent filtering
due to the presence of low mean normalized counts.

Solution: Obtain unfiltered DESeq2 results! (counts without automatic
independent filtering and outlier removal). Then, only the genes with
ALL counts set to zero will have NA for pvalues.

```{r}
dds2_755a <- DESeq(dds_755a, minReplicatesForReplace = Inf)
res2_755a <- results(dds2_755a, 
                           cooksCutoff = FALSE, 
                           independentFiltering = FALSE,
                           contrast = c("ER_microcolonies","HI_LM","LI_LM")
                            )
```

```{r}
head(res2_755a, 30)
```

### Results with thresholds

```{r}
res2df_755a <- as.data.frame(res2_755a) # convert the results table to a df
head(res2df_755a)
```

## MA Plot

```{r}
resultsNames(dds2_755a)
```

```{r}
plotMA_res2_755a <- plotMA(res2_755a, ylim = c(-2, 2))
```

### Histogram of p-values

```{r}
hist(res2_755a$pvalue, breaks = 50, col = "grey50", border = "blue")
```

Further Filtering: baseMean \> 1

```{r}
hist(res2_755a$pvalue[res2_755a$baseMean > 1],
     breaks = 50, col = "grey50", border = "blue")
```

## Annotating and Exporting Results

-   adding gene annotation to results table
-   adding ENTREZ Id to results table

```{r}
res2df_755a$symbol <- mapIds(org.Mm.eg.db,
                                   keys = rownames(dds_755a),
                                  column = "SYMBOL",
                                  keytype = "ENSEMBL",
                                   multiVals = "first")
res2df_755a$entrez <- mapIds(org.Mm.eg.db,
                                   keys = rownames(dds_755a),
                                  column = "ENTREZID",
                                  keytype = "ENSEMBL",
                                   multiVals = "first")
```

```{r}
head(res2df_755a)
```

```{r}
str(res2df_755a)
nrow(res2df_755a)
```

Omit NA values from symbol and respective rows!

```{r}
res3df_755a <- res2df_755a %>% filter(!is.na(symbol) & !is.na(entrez))
```

```{r}
nrow(res3df_755a)
```

## Saving the Results

```{r}
resOrdered_755a <- res3df_755a[order(res3df_755a$pvalue),]
head(resOrdered_755a)
```

```{r}
write.csv(as.data.frame(resOrdered_755a),
          file = "~/R/Alina_RNAseq/755vsLI/results_DGE_EpithelialResponse.csv")
```

## Heatmap of count matrix

To explore a count matrix, it is often instructive to look at it as a
heatmap.

```{r}
select <- order(rowMeans(counts(dds2_755a,normalized = TRUE)),
                decreasing = TRUE)[1:20]
df <- as.data.frame(colData(dds2_755a)[,c("ER_microcolonies","Sample_Name")])
```

```{r}
pheatmap(assay(vsd)[select,], 
         cluster_cols = TRUE, 
         annotation_col = df,
         color = heat.colors,
         show_rownames = FALSE)
```

## Effect of Transformations on Variance

-   These set of plots depict the standard deviation of transformed data
    (across samples), against mean.

### Based on Shifted Log Transformation

```{r}
ntd <- normTransform(dds2_755a)
meanSdPlot(assay(ntd))
```

### Based on Rlog Transformation

```{r}
meanSdPlot(assay(rld))
```

### Based on Variance Stabilizing Transformation

```{r}
meanSdPlot(assay(vsd))
```

## Dispersion Plots

-   Its a useful diagnostic to plot the dispersion estimates.

```{r}
plotDispEsts(dds2_755a)
```

# Volcano Plots

## Volcano Plots based on Enhanced Volcano

```{r fig.height= 10, fig.width= 10}
p1 <- EnhancedVolcano(res3df_755a,
    lab = res3df_755a$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlab = bquote(~Log[2]~'FoldChange'),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    title = 'DE genes: Log2FoldChange Vs -Log10 Pvalue',
    subtitle = bquote(~Log[2]~'|FoldChange| = 1, pvalue < 0.05') ,
    pointSize = 2.0,
    labSize = 5.0,
    boxedLabels = FALSE,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    colAlpha = 0.5,
    legendPosition = 'bottom',
    legendLabSize = 12,
    legendIconSize = 4.0,
    drawConnectors = T,
    widthConnectors = 0.75,
    max.overlaps = 4
    )
(volcano1 <- p1 + scale_y_continuous(limits = c(0, 8),
                                    breaks = seq(0, 8, 1),
                                    sec.axis = sec_axis(~. *1,
                                                        labels = NULL,
                                                        breaks = NULL)) +
                  scale_x_continuous(limits = c(-4, 5), 
                                     breaks = seq(-4, 5, 1),
                                    sec.axis = sec_axis(~. *1,
                                                        labels = NULL,
                                                        breaks = NULL)))

ggsave(filename = '~/R/Alina_RNAseq/755vsLI/Volcano1_L2fcVsPvalue.png',
       plot = volcano1,
       dpi = 300,
       width = 10,
       height = 10,
       units = "in")
```

## volcano Plot of Log2FC vs Padj

```{r fig.height= 10, fig.width= 10}
p2 <- EnhancedVolcano(res3df_755a,
    lab = res3df_755a$symbol,
    x = 'log2FoldChange',
    y = 'padj',
    xlab = bquote(~Log[2]~'FoldChange'),
    ylab = bquote(~Log[10]~'Padj'),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    title = 'DE genes: Log2FoldChange Vs -Log10 Padj',
    subtitle = bquote(~Log[2]~'|FoldChange| = 1, pvalue < 0.05') ,
    pointSize = 2.0,
    labSize = 5.0,
    boxedLabels = FALSE,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    colAlpha = 0.5,
    legendPosition = 'bottom',
    legendLabSize = 12,
    legendIconSize = 4.0,
    drawConnectors = T,
    widthConnectors = 0.75,
    max.overlaps = 4
    )

(volcano2 <- p2 + scale_y_continuous(limits = c(0, 8),
                                    breaks = seq(0, 8, 1),
                                    sec.axis = sec_axis(~. *1,
                                                        labels = NULL,
                                                        breaks = NULL)) +
                  scale_x_continuous(limits = c(-4, 5), 
                                     breaks = seq(-4, 5, 1),
                                    sec.axis = sec_axis(~. *1,
                                                        labels = NULL,
                                                        breaks = NULL)))

ggsave(filename = '~/R/Alina_RNAseq/755vsLI/Volcano2_L2fcVsPadj.png',
       plot = volcano2,
       dpi = 300,
       width = 10,
       height = 10,
       units = "in")
```

## Significant Differentially Expressed Genes

```{r}
res3df_755a$diffexpressed <- "NS"

# if log2Foldchange > 1.0 and pvalue < 0.05, set as "UP"
res3df_755a$diffexpressed[res3df_755a$log2FoldChange > 1.0 & res3df_755a$pvalue < 0.05] <- "UP"

# if log2Foldchange < -1.0 and pvalue < 0.05, set as "DOWN"
res3df_755a$diffexpressed[res3df_755a$log2FoldChange < -1.0 & res3df_755a$pvalue < 0.05] <- "DOWN"

# Create a new column "delabel" to de, that will contain the name of genes differentially expressed (NA in case they are not)
res3df_755a$delabel <- NA
res3df_755a$delabel[res3df_755a$diffexpressed != "NS"] <- res3df_755a$symbol[res3df_755a$diffexpressed != "NS"]
```

Arrive at relevant genes by imposing thresholds.

```{r}
sigsdf_755a <- res3df_755a[(abs(res3df_755a$log2FoldChange) > 1) & (res3df_755a$pvalue < 0.05),]
```

```{r}
nrow(sigsdf_755a)
```

```{r}
summary(res3df_755a)
```

### Gender Genes Removed from Table

```{r}
sigsdf_755a <- filter(sigsdf_755a,
                          symbol != "Xist",
                          symbol != "Jpx", 
                          symbol != "Ftx", 
                          symbol != "Tsx",
                          symbol != "Cnbp2" )

nrow(sigsdf_755a)
#head(sigsdf_755a)
```

Therefore, gender genes werent so much in play in terms of significance!

```{r}

write.csv(sigsdf_755a ,"~/R/Alina_RNAseq/755vsLI/SignificantGenes_ER_mc755A.csv")
```

### Number of Genes from different strains that are contributing to UP/DOWN regulation.

```{r}
sigsdf_755a_UP <- sigsdf_755a[(sigsdf_755a$log2FoldChange) > 1, ] #UP Regulation Table
sigsdf_755a_DOWN <- sigsdf_755a[(sigsdf_755a$log2FoldChange) < -1, ] #DOWN Regulation Table
```

```{r}
nrow(sigsdf_755a_UP)
nrow(sigsdf_755a_DOWN)
```

### Determine Top20 UP Genes and Top20 DOWN Genes

```{r}
UpGene_755a <- sigsdf_755a_UP[order(-sigsdf_755a_UP$log2FoldChange), ]$symbol
DownGene_755a <- sigsdf_755a_DOWN[order(-sigsdf_755a_DOWN$log2FoldChange), ]$symbol
DE_Genes_table_755a <- as.data.frame(cbind(UpGene_755a,DownGene_755a)) # sorted based on highest Log2FC value
```

```{r}
write.csv(DE_Genes_table_755a ,"~/R/Alina_RNAseq/755vsLI/DE_GenesTable_ER_mc755A.csv")
head(DE_Genes_table_755a, 25)
```

```{r}
head(sigsdf_755a)
```

```{r}
# Determining the significant Genes based on Log2FC and pvalue thresholds
sigs2df_755a <- res2df_755a[(abs(res2df_755a$log2FoldChange) > 1) &  (res2df_755a$pvalue < 0.05),]
```

## Z-score based Gene Heatmaps

### with Whole table (all genes together!)

```{r}
mat <- counts(dds2_755a, normalized = TRUE)[rownames(sigsdf_755a),]
mat.zs <- t(apply(mat, 1, scale)) # Calculating the zscore for each row
colnames(mat.zs) <- coldata$Sample_Name # need to provide correct sample names for each of the columns
head(mat.zs)
```

```{r}
mat2 <- counts(dds2_755a, normalized = TRUE)[rownames(sigs2df_755a),]
mat2.zs <- t(apply(mat2, 1, scale)) # Calculating the zscore for each row
colnames(mat2.zs) <- coldata$Sample_Name # need to provide correct sample names for each of the columns
head(mat2.zs)
#pheatmap(mat2.zs, cluster_cols = TRUE, cluster_rows = FALSE, show_rownames = FALSE)
```

```{r fig.height= 15, fig.width= 15}
newHP <- Heatmap(mat2.zs, 
        cluster_columns = TRUE, 
        cluster_rows = TRUE, 
        column_labels = colnames(mat2.zs),
        name = 'DE Genes Heatmap',
        show_row_names = FALSE, 
        use_raster = TRUE, 
        raster_quality = 5,
        #width = unit(8, "in"), 
        #height = unit(8, "in")
        row_labels = sigs2df_755a[rownames(mat2.zs),]$symbol
        )
newHP
```

```{r fig.height= 140, fig.width= 10}
LongHeatMap_Allgenes <- Heatmap(mat2.zs, 
        cluster_columns = TRUE, 
        cluster_rows = TRUE, 
        column_labels = colnames(mat2.zs),
        row_labels = sigs2df_755a[rownames(mat2.zs),]$symbol,
        name = 'Z-Score Heatmap of DE Genes - ER_microcolinies_755A',
        show_row_names = TRUE, 
        use_raster = TRUE, 
        raster_quality = 5,
        )
LongHeatMap_Allgenes
#saveplot(LongHeatMap_Allgenes, "LongHeatMap_Allgenes")
```

Need to filter these results to accommodate better the heat maps and
also volcano plots

### with tighter constraints (all genes together!)

```{r}
sigs1df_755a <- res2df_755a[(res2df_755a$baseMean > 75) & (abs(res2df_755a$log2FoldChange) > 2) & (res2df_755a$pvalue < 0.05),]
mat1 <- counts(dds2_755a, normalized = TRUE)[rownames(sigs1df_755a),]
mat1.zs <- t(apply(mat1, 1, scale)) # Calculating the zscore for each row
colnames(mat1.zs) <- coldata$Sample_Name # need to provide correct sample names for each of the columns
head(mat1.zs)
```

```{r fig.height= 10, fig.width= 10}
Heatmap(mat1.zs,
        cluster_columns = TRUE,
        cluster_rows = TRUE,
        column_labels = colnames(mat1.zs),
        name = 'DE Genes',
        row_labels = sigs1df_755a[rownames(mat1.zs),]$symbol)
```

# GO Terms with clusterProfiler

## GO Terms for UP Regulated Genes

### GO over-representation analysis for UP Regulated Genes

```{r}
UPgene_ENS_ID <- rownames(sigsdf_755a_UP)

GO_UPRegResults_755a <- enrichGO(gene = UPgene_ENS_ID,
                                       OrgDb = "org.Mm.eg.db",
                                       keyType = "ENSEMBL",
                                       ont = "BP",
                                       pAdjustMethod = "BH",
                                       pvalueCutoff  = 0.01,
                                       qvalueCutoff  = 0.05,
                                       readable      = TRUE)
```

```{r}
GO_UpRegdf_755a <- as.data.frame(GO_UPRegResults_755a)
head(GO_UpRegdf_755a)
```

```{r}
write.csv(GO_UpRegdf_755a ,"~/R/Alina_RNAseq/755vsLI/GO_UpReg_results_ERMC_755A.csv")
```

```{r fig.height=10, fig.width=10}
GO_UPReg_Barplot_755a <- plot(barplot(GO_UPRegResults_755a, showCategory = 22))
saveplot(GO_UPReg_Barplot_755a, "GO_UPReg_Barplot_755a")
```

```{r fig.height=10, fig.width=10}
GO_UPReg_Dotplot_755a <- plot(dotplot(GO_UPRegResults_755a, showCategory = 23))

saveplot(GO_UPReg_Dotplot_755a, "GO_UPReg_Dotplot_755a")
```

```{r fig.height=12, fig.width=10}
GO_UPReg_Cnetplot_755a <- plot(cnetplot(GO_UPRegResults_755a, showCategory = 13))

saveplot(GO_UPReg_Cnetplot_755a, "GO_UPReg_Cnetplot_755a")
```

### Upset Plot

```{r fig.height= 8, fig.width=12}
GO_UPReg_Upsetplot <- plot( upsetplot(GO_UPRegResults_755a))
saveplot(GO_UPReg_Upsetplot, "GO_UPReg_Upsetplot")
```

### Heatplot

The heatplot is similar to cnetplot, while displaying the relationships
as a heatmap. The gene-concept network may become too complicated if
user want to show a large number significant terms. The heatplot can
simplify the result and more easy to identify expression patterns.

```{r fig.height= 15, fig.width=15}
GO_UPReg_Heatplot <- plot( heatplot(GO_UPRegResults_755a))
saveplot(GO_UPReg_Heatplot, "GO_UPReg_Heatplot")
```

### Tree Plot of Enriched Terms

```{r fig.height= 10, fig.width=15}
edox2 <- pairwise_termsim(GO_UPRegResults_755a)
GO_UPReg_enrichtreeplot <- plot(treeplot(edox2))
saveplot(GO_UPReg_enrichtreeplot, "GO_UPReg_enrichtreeplot")
```

## GO Terms for Down Regulated Genes

### GO over-representation analysis for DOWN Regulated Genes

```{r}
DOWNgene_ENS_ID <- rownames(sigsdf_755a_DOWN)
```

```{r}
GO_DOWNRegResults_755a <- enrichGO(gene = DOWNgene_ENS_ID,
                       OrgDb = "org.Mm.eg.db",
                       keyType = "ENSEMBL",
                       ont = "BP",
                       pAdjustMethod = "BH",
                       pvalueCutoff  = 0.01,
                       qvalueCutoff  = 0.05,
                       readable      = TRUE
                       )
GO_DOWNRegdf_755a <- as.data.frame(GO_DOWNRegResults_755a)
head(GO_DOWNRegdf_755a)
```

```{r}
write.csv(GO_DOWNRegdf_755a,
          "~/R/Alina_RNAseq/755vsLI/GO_DOWNReg_results_ERMC_755A.csv")
```

```{r fig.height= 10, fig.width=10}
GO_DOWNReg_Barplot_755a <- plot(barplot(GO_DOWNRegResults_755a, showCategory = 22))
saveplot(GO_DOWNReg_Barplot_755a, "GO_DOWNReg_Barplot_755a")
```

```{r fig.height=10, fig.width=10}
GO_DOWNReg_Dotplot_755a <- plot(dotplot(GO_DOWNRegResults_755a, showCategory = 23))

saveplot(GO_DOWNReg_Dotplot_755a, "GO_DOWNReg_Dotplot_755a")
```

```{r fig.height= 10, fig.width=10}
GO_DOWNReg_Cnetplot_755a <- plot(cnetplot(GO_DOWNRegResults_755a, 
                                                showCategory = 13
                                                ))

saveplot(GO_DOWNReg_Cnetplot_755a, "GO_DOWNReg_Cnetplot_755a")
```

### Upset Plot

The upsetplot is an alternative to cnetplot for visualizing the complex
association between genes and gene sets. It emphasizes the gene
overlapping among different gene sets.

```{r fig.height= 8, fig.width=15}
GO_DOWNReg_Upsetplot <- plot( upsetplot(GO_DOWNRegResults_755a))
saveplot(GO_DOWNReg_Upsetplot, "GO_DOWNReg_Upsetplot")
```

### Heat Plot

```{r fig.height= 12, fig.width=20}
GO_DOWNReg_Heatplot <- plot(heatplot(GO_DOWNRegResults_755a))
saveplot(GO_DOWNReg_Heatplot, "GO_DOWNReg_Heatplot")
```

### Tree Plot of Enriched Terms

```{r fig.height= 10, fig.width=15}
edox1 <- pairwise_termsim(GO_DOWNRegResults_755a)
GO_DOWNReg_enrichtreeplot <- plot(treeplot(edox1))
saveplot(GO_DOWNReg_enrichtreeplot, "GO_DOWNReg_enrichtreeplot")
```

## Pathway Analysis

### KEGG Pathways

1.  The gageData package has pre-compiled databases mapping genes to
    KEGG pathways and GO terms for common organisms.
2.  kegg.sets.mm is a named list. Each element is a character vector of
    member gene Entrez IDs for a single KEGG pathway. sigmet.idx.hs is
    an index of numbers of sinaling and metabolic pathways in
    kegg.set.gs.
3.  In other words, KEGG pathway include other types of pathway
    definitions, like "Global Map" and "Human Diseases", which may be
    undesirable in pathway analysis. Therefore,
    kegg.sets.mm[sigmet.idx.mm] gives the "cleaner" gene sets of
    sinaling and metabolic pathways only.

The gage() function requires a named vector of fold changes, where the
names of the values are the Entrez gene IDs.

clusterProfiler requires the genes to be identified using Entrez IDs for
all genes in our results dataset. We also need to remove the NA values
and duplicates (due to gene ID conversion) prior to the analysis

```{r}
## Remove any NA values (reduces the data by quite a bit) and Remove any Entrez duplicates
entreztable <- unique(dplyr::filter(sigsdf_755a, entrez != "NA"))
```

extract and name the fold changes

```{r}
foldchanges <-  sigsdf_755a$log2FoldChange
names(foldchanges) <-  sigsdf_755a$entrez

## Sort fold changes in decreasing order
foldchanges <- sort(foldchanges, decreasing = TRUE)
head(foldchanges)
```

```{r}
data(kegg.sets.mm)
data(sigmet.idx.mm)
kegg.sets.mm <-  kegg.sets.mm[sigmet.idx.mm]
```

Here, I am using same.dir = TRUE, which will give us separate lists for
pathways that are upregulated versus pathways that are downregulated.
Let's look at the first few results from each.

```{r}
# Get the results
keggres = gage(exprs = foldchanges, 
               gsets = kegg.sets.mm, 
               same.dir = TRUE)
```

```{r}
# Look at both up (greater), down (less), and statatistics.
lapply(keggres, head)

```

Now, process the results to pull out the top 20 up-regulated pathways,
then further process that just to get the IDs. I can use these KEGG
pathway IDs downstream for plotting.

```{r}
# Get the pathways
# Get the pathways
keggrespathways = data.frame(id = rownames(keggres$greater),
                             keggres$greater) %>% 
                  tibble::as_tibble() %>% 
                  filter(row_number() <= 10) %>% 
                  .$id %>% 
                  as.character()
keggrespathways
```

```{r}
# Get the IDs.
keggresids = substr(keggrespathways, start = 1, stop = 8)
keggresids
```

The pathview() function in the pathview package makes the plots. The
function is written below to loop through and draw plots for the top 10
pathways we created above.

```{r}
# Define plotting function for applying later
plot_pathway = function(pid) pathview(gene.data = foldchanges, 
                                      pathway.id = pid, 
                                      species = "mouse", 
                                      new.signature = FALSE)


# plot multiple pathways (plots saved to disk and returns a throwaway list object)
tmp = sapply(keggresids, function(pid) pathview(gene.data = foldchanges, 
                                                pathway.id = pid, 
                                                species = "mouse"))
```
## Signaling Pathway Impact Analysis

Preparing different Paramters for SPIA:

Note(from Vignette): The DE_Colorectal is a vector containing the log2 fold changes of the genes found to be differentially expressed between cancer and normal samples, and ALL_Colorectal is a vector with the Entrez IDs of all genes profled on the microarray. The names of the DE_Colorectal are the Entrez gene IDs corresponding to the computed log fold-changes.

1. de must be a vector of log2 fold changes. The names of de should be included in the refference array!

```{r}
de_l2fc <- as.vector(res3df_755a$log2FoldChange)
# <- res3df_755a$entrez <-
#e_ID <- res3df_755a$entrez
head(de_l2fc)
```

```{r}
#de_l2fc_df <- as.data.frame(de_l2fc)
#de_l2fc_df
```


```{r}
backgroundgenes <- !is.na(sigsdf_755a$entrez)
#backgroundgenes <- backgroundgenes[((backgroundgenes))]

```

```{r}
#spia_result <- spia(de = as.vector(de_l2fc),all = backgroundgenes, organism = "mmu")
```
```{r}
sessionInfo()
```

## GSEA - Gene Set Enrichment Analysis

```{r}
head(foldchanges)
```

Perform the GSEA using KEGG gene sets:

```{r}
library("stats")
## GSEA using gene sets from KEGG pathways
gseaKEGG <- gseKEGG(geneList = foldchanges, # ordered named vector of fold changes (Entrez IDs are the associated names)
              organism = "mmu", # mmu is Mus Musculus
              #nPerm = 1000, # default number permutations
              minGSSize = 20, # minimum gene set size (# genes in set) - change to test more sets or recover sets with fewer # genes
              pvalueCutoff = 0.05, # padj cutoff value
              verbose = T)

## Extract the GSEA results
gseaKEGG_results <- gseaKEGG@result
```

How many pathways are enriched? View the enriched pathways:

## Write GSEA results to file

```{r}
head(gseaKEGG_results)

write.csv(gseaKEGG_results, "~/R/Alina_RNAseq/755vsLI/gseaOE_kegg.csv")
```

Explore the GSEA plot of enrichment of one of the pathways in the ranked
list:

```{r fig.width=10, fig.height=6}
## Plot the GSEA plot for a single enriched pathway, `mmu04621`
gseaplot(gseaKEGG, geneSetID = 'mmu04972')
```

Use the Pathview R package to integrate the KEGG pathway data from
clusterProfiler into pathway images:

```{r}
detach("package:dplyr", unload = TRUE) # first unload dplyr to avoid conflicts

## Output images for a single significant KEGG pathway
pathview(gene.data = foldchanges,
              pathway.id = "mmu04972",
              species = "mmu",
              limit = list(gene = 2, # value gives the max/min limit for foldchanges
              cpd = 1))
```

# Gene Ontology - Hyper Geometric Test

```{r}
selectUPGenes_entrezid = unique(sigsdf_755a_UP['entrez'] )
selectDOWNGenes_entrezid = unique(sigsdf_755a_DOWN['entrez'] )
```

```{r}
UniverseGenes = unique(sigsdf_755a$entrez)
cutOff = 0.01 # setting the cutoff at 1%
```

```{r}
upParams = new("GOHyperGParams",
               geneIds = selectUPGenes_entrezid,
               universeGeneIds = UniverseGenes,
               annotation = "org.Mm.eg.db",
               ontology = "BP",
               pvalueCutoff = cutOff,
               testDirection = "over")
downParams = new("GOHyperGParams",
               geneIds = selectDOWNGenes_entrezid,
               universeGeneIds = UniverseGenes,
               annotation = "org.Mm.eg.db",
               ontology = "BP",
               pvalueCutoff = cutOff,
               testDirection = "over")
```

```{r}
upBP = hyperGTest(upParams)
ubp1 <- as.data.frame(summary(upBP))
ubp1
```
```{r fig.width=15, fig.height=10}
plot(ubp1)
```

```{r}
downBP = hyperGTest(downParams)

```
```{r}
dbp1 <- as.data.frame(summary(downBP))
dbp1
```

```{r fig.width=15, fig.height=10}
plot(dbp1)
```
```{r}
htmlReport(downBP, file = "downBP.html")
```

