---
title: "Lets Go: From Bams to Count Matrix Notebook"
output: html_notebook
---

```{r}
library(Rsubread)
```

# Feature counts
```{r}
# GTF File
gtf_file <- "C:/Users/User/Documents/bam/Mus_musculus.GRCm39.106.chr.gtf"

```

```{r}
# Bam files
bam_path <- "C:/Users/User/Documents/bam/"
bamFiles <- list.files(bam_path, pattern = ".*bam$")
bamFiles
```


```{r}
setwd(bam_path)
fc <- featureCounts(files = bamFiles,
                    annot.ext= gtf_file,
                    isGTFAnnotationFile=TRUE,
                    isPairedEnd=FALSE)
```



```{r}
head(fc)
```
```{r}
head(fc[["annotation"]], 10)

```

```{r}
head(fc[["counts"]], 10)
write.csv(fc[["counts"]], 'C:/Users/User/Documents/bam/counts.csv')
```


# Differential Gene Expression Analysis

```{r}
library("DESeq2")
library("PoiClaClu")
library("pheatmap")
library("RColorBrewer")
library('tidyverse')
library("PoiClaClu")
library("vsn")
library('EnhancedVolcano')
library('gplots')
library('org.Mm.eg.db')
library('stringr')
```



## Creating metadata for the DGE Analysis

DESeq2 needs sample information (metadata) for performing DGE analysis. Let’s create the sample information 
```{r}
# write the list of bam files as a list to be saved in csv file
write.csv(list(bamFiles), 'C:/Users/User/Documents/bam/samples.csv')
```



```{r}
# Read the csv file and change the column name 
sample_ID <- read.csv('C:/Users/User/Documents/bam/samples.csv')
names(sample_ID)[2]  <- 'Sample_Name'
colnames(sample_ID)
```


```{r}
condition = c("Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","Infected","control","control")
```


```{r}
coldata <- data.frame(sample_ID, condition)
colnames(coldata) <- c('Number','Sample_Name','condition') # change name of one of the columns
coldata <- subset (coldata, select = -Number)
```

the metadata can be foudn in a df called coldata!
```{r}
head(coldata)
```

### Tidying up the names for plots later!

####First from coldata
```{r}

coldata$Sample_Name = str_remove_all(coldata$Sample_Name, pattern = 'run6_trimmed_') # tidying up the names
#coldata$condition <- as.factor(coldata$condition)
coldata <- data.frame(coldata, row.names = 1) # convert column to row.names
coldata 
```

#### then fix Countsmatrix:

NOTE:

1. From the manuals the countsData must be a numeric matrix
2. It is IMPORTANT to keep the names of the genes in the rownames


```{r}
countsmatrix <-as.matrix(read.csv('C:/Users/User/Documents/bam/counts.csv'))
```

```{r}
dim(countsmatrix)
str(countsmatrix)
```

```{r}
colnames(countsmatrix) = str_remove_all(colnames(countsmatrix), pattern = 'run6_trimmed_') #tidying up these names again
colnames(countsmatrix)
```

```{r}
rownames(countsmatrix) <- countsmatrix[,1] #converting first column into rownames for sanity check later
# It is IMPORTANT to keep the names of the genes in the rownames

```


```{r}
countsmatrix<- subset (countsmatrix, select = -X)#dropping the X column
colnames(countsmatrix)
```


```{r}

class(countsmatrix) <- "numeric"
str(countsmatrix)
dim(countsmatrix)
```


## DESeq2 Analysis


Now, construct DESeqDataSet for DGE analysis:


But before that, a sanity check :

It is essential to have the name of the columns in the count matrix in the same order as that in name of the samples (rownames in coldata).

```{r}
all(rownames(coldata) %in% colnames(countsmatrix))
```

```{r}
ncol(countsmatrix) == nrow(coldata)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = countsmatrix, 
                              colData = coldata, 
                              design = ~ condition)
```

```{r}
dds
```



